---
output: html_document
---
## Assessment of severe weather impact using data in the NOAA Storm Database"
  
---
  
#### Reproducible Research: Peer Assessment 2
#### Daniel Brown
#### 31/08/2020
  
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

### Synopsis
```{r synopsis, include = FALSE, echo = FALSE, ref.label = c("loadData", "convDates", "deaths", "injuries")}
x <- allDeaths
```
Total deaths is `r allDeaths`  
  
### Data Processing

Data is expected in a file called *repdata_data_StormData.csv* in the working directory.

Read in the data and load it to a data frame named *sData*:

```{r loadData, cache = TRUE}
zipfilePath <- paste("stormData.zip", sep = "")
         url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
         if (!file.exists("./repdata_data_StormData.csv")) {
                 if (!file.exists(zipfilePath)) {
                         download.file(url, zipfilePath)
                 }
                 unzip(zipfilePath, overwrite = TRUE)
         }

sData <- read.csv("repdata_data_StormData.csv", header = TRUE, sep = ",")
```
  
Convert the start/end date fields so they conform with POSIXct standards and can be used in calculations.
  
```{r convDates}
sData$BGN_DATE <- strptime(sData$BGN_DATE, "%m/%d/%Y %H:%M:%S")
sData$END_DATE <- strptime(sData$END_DATE, "%m/%d/%Y %H:%M:%S")
```

### Results
##### 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Impact to population health will be gauged by reference to the following variables:  
  
* FATALITIES: The number of deaths attributed to a storm event; or  
* INJURIES: The number of injuries attributed to a storm event.  
  
This process derives a set of data to show us the 10 *EVTYPE* values that are associated to the highest number of deaths, as recorded in the *FATALITIES* field.

```{r deaths}
library(dplyr)
library(ggplot2)

allDeaths <- sum(sData$FATALITIES, na.rm = TRUE)

tDeaths <- sData %>% 
                group_by(EVTYPE) %>%
                summarise(tDeaths = sum(FATALITIES)) %>%
                arrange(desc(tDeaths))
mostDeaths <- head(tDeaths, 10)
mostDeaths$pc <- mostDeaths$tDeaths / allDeaths

ggplot(mostDeaths, aes(x = reorder(EVTYPE, desc(tDeaths)), y = tDeaths, fill = EVTYPE)) +
        geom_bar(stat = 'identity', position = "identity") +
        guides(fill = FALSE) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
        xlab(NULL) +
        ylab("Total Deaths") +
        labs(title = "Top 10 causes of death by storm event ('EVTYPE')")
```
  
The top 10 events account for `r sprintf("%.1f%%", 100*sum(mostDeaths$pc))` of all recorded deaths in the dataset.
  
Then we can derive a set of data to show us the 10 *EVTYPE* values that are associated to the highest number of deaths, as recorded in the *INJURIES* field.
  
```{r injuries}
allInjuries <- sum(sData$INJURIES, na.rm = TRUE)

tInjuries <- sData %>% 
                group_by(EVTYPE) %>%
                summarise(tInjuries = sum(INJURIES)) %>%
                arrange(desc(tInjuries))
mostInjuries <- head(tInjuries, 10)
mostInjuries$pc <- mostInjuries$tInjuries / allInjuries

ggplot(mostInjuries, aes(x = reorder(EVTYPE, desc(tInjuries)), y = tInjuries, fill = EVTYPE)) +
        geom_bar(stat = 'identity', position = "identity") +
        guides(fill = FALSE) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
        xlab(NULL) +
        ylab("Total Injuries") +
        labs(title = "Top 10 causes of injury by storm event ('EVTYPE')")
```

The top 10 events account for `r sprintf("%.1f%%", 100*sum(mostInjuries$pc))` of all recorded injuries in the dataset.

##### 2. Across the United States, which types of events have the greatest economic consequences?

Results here